---
title: "Project 1"
author: "Adam Douglas"
date: "9/16/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Text File

First we read in the raw chess text file. By using the `read_table` method from the `readr` package gets us a single long character field for each row:

```{r load file, warning=FALSE, message=FALSE, results='hide'}
chess <- read_table("chess.txt",col_names="txt")
```
```{r}
head(chess)
```

## Data Prep

Before we can extract the actual data we are looking to record, we need to do a few steps to prepare the raw text data:

```{r prepare text}
# Filter to the rows with data only by looking for the pipe character
chess <- chess %>% filter(str_detect(chess$txt,"\\|"))

# Eliminate the first two header rows
chess <- chess[3:length(chess$txt),]

# Split the lines by pipe "|" characters
temp <- str_split(chess$txt,"\\|")
```

Now that we have a cleaner dataset, we now have to parse through and get the specific data we're after. First we parse the odd rows:

```{r odd rows}
# We parse the odd rows to get: number, name, points, and opponent numbers

# Set-up the initial vectors, by initializing with dummy data
number <- -999
names(number) <- "number"

name <- "DUMMY DATA"
names(name) <- "name"

point <- 0.0
names(point) <- "points"

opponent <- data.frame(rbind(c(0,0,0,0,0,0,0)))
names(opponent) <- c("opp1","opp2","opp3","opp4","opp5","opp6","opp7")

# Iterate through each odd line
for(i in seq(1,length(temp),2)){
  x <- temp[[i]][[1]] %>% str_trim() %>% parse_number() # number
  n <- temp[[i]][[2]] %>% str_trim() # name
  p <- temp[[i]][[3]] %>% str_trim() %>% parse_number() # points
  s <- data.frame(rbind(str_extract(temp[[i]][4:10],"[[:digit:]]+") %>% parse_number())) # opponents
  names(s) <- c("opp1","opp2","opp3","opp4","opp5","opp6","opp7")
  
  number <- rbind(number, x)
  name <- rbind(name,n)
  point <- rbind(point,p)
  opponent <- rbind(opponent,s)
}
```

Next, the even rows:

```{r even rows}
# We parse the even rows to get: state, and rating

# Set-up the initial vectors, by initializing with dummy data
state <- "XX"
names(state) <- "state"

rating <- 0
names(rating) <- "rating"

# Iterate through each even line
for(i in seq(2,length(temp),2)){
  t <- temp[[i]][[1]] %>% str_trim() # state
  r <- temp[[i]][[2]] %>% str_extract("(?<=R:\\s{0,4})[[:digit:]]+") %>% parse_number() # rating
  
  state <- rbind(state,t)
  rating <- rbind(rating,r)
}
```

Now we put it all together:

```{r combine data}
# Combine all the elements
data <- data.frame(number,name,state,rating,point,opponent,stringsAsFactors = FALSE)
row.names(data) <- data$number
data <- filter(data, number != -999) # Remove our dummy row
```

Finally, we do our opponent rating lookups:
```{r lookup ratings}
# A helper function to get a player's rating
getRating <- function(n){
  if(is.na(n)) return(NA)
  data %>% filter(number == n) %>% select(rating)
}

# Get the ratings into a new data frame
opp <- data.frame(opp1 = unlist(lapply(data$opp1,getRating), use.names = FALSE),
                 opp2 = unlist(lapply(data$opp2,getRating), use.names = FALSE),
                 opp3 = unlist(lapply(data$opp3,getRating), use.names = FALSE),
                 opp4 = unlist(lapply(data$opp4,getRating), use.names = FALSE),
                 opp5 = unlist(lapply(data$opp5,getRating), use.names = FALSE),
                 opp6 = unlist(lapply(data$opp6,getRating), use.names = FALSE),
                 opp7 = unlist(lapply(data$opp7,getRating), use.names = FALSE)
                 )
# Fine the mean opponent rating
data$oppMean <- round(apply(opp,1,mean, na.rm=TRUE),0)
```

## Export

We have all of the data elements we need and can now export to a CSV file:

```{r, eval=FALSE}
# Select and export our fields
data %>% select(name,state,points,rating,oppMean) %>% write_csv("data.csv", col_names=FALSE)
```

