---
title: "Week 7 Assignment"
author: "Adam Douglas"
date: "10/13/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RCurl)
library(XML)
library(RJSONIO)
```

## The Individual Books

I am an avid reader and have been for many years. I've always tended towards the fantasy genre, so my choices of "favorites" reflect that preference. Some of these choices I read early in life, some more recently, but each one I have read at least a few times and enjoy them all over again each time I do.

### First Book

The first, and probably most influential book, I loaded into an HTML table. With the XML package, I can retreive it and parse it into a data frame:

```{r book1}
# GitHub location
loc <- "https://raw.githubusercontent.com/lysanthus/Data607/master/Week7/book1.html"

# Get the HTML
html <- getURL(loc)

# Parse the HTML
book <- htmlParse(html)

# Get the table headers as column names
headers <- xpathSApply(book,"//th",xmlValue)

# Get the table data for actual book info
data <- xpathSApply(book,"//td",xmlValue)

# Combine into a data frame. Note we had to change the data to a single row.
#   If there had been more than one book, we may have had to take a different
#   approach.
book1 <- data.frame(rbind(data), stringsAsFactors = FALSE, row.names="Book1")

colnames(book1) <- headers

book1
```

This book, `r book1[1,1]`, was more or less my gateway into the genre of fantasy fiction. I still like to revisit this (and the others in its trilogy) every few years. In, fact, I may be due for another read.

### Second Book

For my second book, I chose a true classic - a book likely on the shelves of many, even those not particularly enamoured with the fantasy genre.

This book, I loaded into a simple XML file.

```{r book2}
# GitHub Location
loc <- "https://raw.githubusercontent.com/lysanthus/Data607/master/Week7/book2.xml"

# Get the XML
xml <- getURL(loc)

# Parse the HTML
book <- xmlParse(xml)

# Get the fields
data <- xpathSApply(book,"//book/child::*",xmlValue)

# The column names
headers <- c("title","author","published","pages")

book2 <- data.frame(rbind(data), stringsAsFactors = FALSE, row.names="Book2")

colnames(book2) <- headers

book2

```

This book, `r book2$title` is a classic and a must-read for any serious fan of fantasy literature.

### Third Book

For the third book, I chose the first of a trilogy I only recently read. Few books developed characters quite as engaging as this did, and the world the author created was rather unique from other novels.

This book I placed into a JSON file.

```{r book3}
# GitHub Location
loc <- "https://raw.githubusercontent.com/lysanthus/Data607/master/Week7/book3.json"

# Get the JSON
json <- getURL(loc)

# Parse the JSON
book <- fromJSON(content=json)

# Get the fields
data <- unlist(book, recursive = TRUE)

# The column names
headers <- c("title","author","published","pages")

book3 <- data.frame(rbind(data), stringsAsFactors = FALSE, row.names="Book3")

colnames(book3) <- headers

book3
```

When I discovered `r book3$author`, I was surprised that I had not heard of her books before. They instantly drew me in and captured my imagination throughout the two trilogies.

## Analysis

HTML and XML use similar approaches to parsing, while JSON is slightly different. Neither is particularly difficult to do (though I suspect multiple entries in each are a bit trickier to coax into a data frame).

A far as the individual data frames, each are slightly different in that they have different column names (which I could have manually fixed), but otherwise the same:

```{r}
str(book1)
str(book2)
str(book3)
```

Each variable is imported as a character, which is ok for now and can be coerced into its proper type later if we were to combine them into a single dataframe:

```{r allbooks}
# Fix column names for book1
names(book1) <- c("title","author","published","pages")

# Bind into a single data frame
books <- rbind(book1,book2,book3)

# Fix columns
books$published <- as.numeric(books$published)
books$pages <- as.numeric(books$pages)

books
```

***
