---
title: "Project 3 Twitter Mining"
output:
  pdf_document: default
  html_document: default
---
#Access variables for twitter (commented out)
```{r}
#ckey <- ""
#secret_key <-""
#ctoken <- ""
#secret_token <- ""
```

#Install the necessary packages
```{r}
library(tm)
library(twitteR)
library(wordcloud)
library(RColorBrewer)
library(e1071)
library(class)
```

#Set up the twitter connection from account
```{r}
setup_twitter_oauth(ckey,secret_key,ctoken,secret_token)
```
#List of words we're searching for in tweets
```{r}
data_science_list <- c("datascience", "data science", "data", "science", "scientist", "datascientist")
```
# pull n tweets from twitter
```{r}
data.science.tweets <- searchTwitter(data_science_list, n = 100000, lang = 'en')
```
# get text data
```{r}
data.science.text <- sapply(data.science.tweets, function(x) x$getText())
```
# clean text data

# remove emojis etc - turns to UTF-8
```{r}
data.science.text <- iconv(data.science.text, 'UTF-8', 'ASCII')
```
# Create corpus
```{r}
ds.corpus <- Corpus(VectorSource(data.science.text))
```
# doc term matrix, includes stop words to not include in matrix
```{r}
term.doc.matrix <- TermDocumentMatrix(ds.corpus,
                                      control = list(removePunctuation = TRUE,
                                                     stopwords = c("datascience", "data science", "data", "science", "scientist", "datascientist","datascientists", "right", "now", "abc", "via", "back","words", "lifehacks", "can", "big", "nice", "users","free",'using','good','mit','new','need','art','build','great','cool','next','https','http', 
                                                                   stopwords('english')),
                                                                   removeNumbers = TRUE,
                                                                   tolower = TRUE))

```
# Convert into matrix
```{r}
term.doc.matrix <- as.matrix(term.doc.matrix)
```

# word counts in decreasing order
```{r}
word.freq <- sort(rowSums(term.doc.matrix), decreasing = TRUE)
dm <- data.frame(word = names(word.freq), freq = word.freq)
```
# Select top 25 words
```{r}
dm <- dm[1:25,]
```
# word cloud
```{r}
wordcloud(dm$word,dm$freq, random.order = FALSE, colors = brewer.pal(8,'Dark2'))
```

