---
title: "RSS Proof of Concept"
author: "Adam Douglas"
date: "10/11/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RCurl)
library(XML)
library(tidyverse)
library(wordcloud)
```

## Outline

If we can get a decent sample of blogs and read their RSS feeds, we can possibly mine them for category tags and filter that as a list of skills. The idea is that bloggers write about topics of most interest in Data Science (of which Data Science skills are a subset).

### KDNuggets

The only real Data Science blog I am familiar with is [KDNuggets](https://www.kdnuggets.com/news/index.html). They have an RSS feed [here](https://www.kdnuggets.com/feed).

As a proof of concept, this is what I did:

```{r}
# Use RCurl to get the raw RSS/XML (XML package won't read https for some reason)
kd <- getURL("https://www.kdnuggets.com/feed")

# Save as a file
write_file(kd,"kd.rss")

# Parse the XML
test <- xmlParse("kd.rss")

# Use XPath to get the category
categories <- data.frame(table(xpathSApply(test,"//category", xmlValue)))

# Look at the categories
categories %>% arrange(desc(Freq))
```

## Words

```{r}
filtered <- categories %>% filter(!str_detect(Var1,"2018"),!str_detect(Var1,"Data Science"),Var1 != "Jobs", Var1 != "Online Education") %>% arrange(desc(Freq))

filterd <- filtered[1:25,]

wordcloud(filtered$Var1, filtered$Freq, colors = brewer.pal(8,'Dark2'),scale=c(2,.25))
```

### Others?
I am sure this can be repeated for other, legitimate DS blogs.
