---
title: "DATA 607 - Week 3 Assignment"
author: "Adam Douglas"
date: "9/11/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
```

## Problems

**3. Copy the introductory example. The vector `names` stores the extracted names.**

```{r}
# Introductory text

raw.data <-"555-1239Moe Szyslak(636) 555-0113Burns, C. Montgomery555-6542Rev. Timothy Lovejoy555 8904Ned Flanders636-555-3226Simpson, Homer5553642Dr. Julius Hibbert"

name <- unlist(str_extract_all(raw.data, "[[:alpha:]., ]{2,}"))

name
```

(a) Use the tools of this chapter to rearrange the vector so that all elements conform to the standard `first_name last_name`.

```{r}
# Because we have two formats: First Last and Last, First
#   the first name is either a word terminated by a space (including 
#   a single initial with a period) or
#   a word at the end of the string.

first <- str_trim(str_extract(name,"(\\w+ )|(\\w{1}\\. )?+(\\w+$)"))

first

# So, the last name is either the word before the comma or
#   the word at the end of the string

last <- str_extract(name,"(\\w+,)|(\\w+$)")

last

# Now we can combine into one vector
firstLast <- str_c(first," ",last)

firstLast
```

(b) Construct a logical vector indicating whether a character has a title (i.e. Rev. and Dr.)

```{r}
# For a title, we look for 2 or more letters and a period

title <- str_detect(name,"\\w{2,}\\.")

title
```

(c) Construct a logical vector indicating whether a character has a second name.

```{r}
# To locate whether someone has a second name, we look for a period in the
#   first name from the initial

twoNames <- str_detect(first,"\\.")

twoNames
```

***

**4. Describe the types of strings that conform to the following regular expressions and construct an example that is matched by the regular expression.**

(a) `[0-9]+\\$`
    
    This pattern matches one or more digits between 0 and 9 and a dollar sign. One example that would match: "129$"

(b) `\\b[a-z]{1,4}\\b`
    
    This pattern matches words that are any combination of 1,2,3, or 4 lower-case letters. An example of a match here would be "adam"

(c) `.*?\\.txt$`
    
    Text that matches this pattern would contain any character (or no characters!) with the string ".txt" at the end. The text "filename.txt" would be matched by this, for example.

(d) `\\d{2}/\\d{2}/\\d{4}`

    This pattern matches 2 digits, a slash, 2 more digits, another slash, and 4 digits. This matches date strings such as "11/30/2017".

(e) `<(.+?)>.+?</\\1>`

    This regex matches HTML tags such as "<tr>Row Title</tr>" since it looks for angle brackets with any characters inside, characters after that, and the same angle-bracket text with a / added to the beginning (closing tag).

***
